--- 
title: 01. Intro to data viz
execute:
  echo: false
  warning: false
  message: false
--- 

## What is data visualization?

It's pretty common to find a book with hundreds of pages of details on data
visualization, but no definition. Two I've found useful:

> The rendering of information in a visual format to help communicate data while
> also generating new patterns and knowledge through the act of visualization
> itself [@wilson2018 p. 8]

> The representation and presentation of data to facilitate understanding
> [@kirk2016 p. 19]

## What should visualization do?

> Data visualization is part art and part science. The challenge is to get the
> art right without getting the science wrong and vice versa. A data
> visualization first and foremost has to accurately convey the data. It must
> not mislead or distort.... At the same time, a data visualization should be
> aesthetically pleasing [@wilke2019, ch 1]

## Why visualize data?

As you create data viz, you'll have different purposes. I think of them in two
big categories:

- Visualization for myself / my colleagues (i.e. private)
- Visualization for others (i.e. public)

In order to make a public chart, I make many, many private ones.

@kirk2016 uses an interesting framework (p. 76) where visualization can be
mapped along two axes, experience and tone. Tone is broken into feeling vs
reading, and experience is broken into explaining, exhibiting, and exploring.
We'll come back to this framework when we think about the purposes of our
visualizations, what we want to convey, and how.

## How can data visualization be used?

::: {.callout-tip}
### Brainstorming

#### Positive / constructive

- predicting weather
- making preparations
- sports analytics
- awareness in health
- explaining / understanding scope
- making an argument
- mitigating risk
- communicating risk / possible outcomes / dangers

#### Negative / destructive

- skewed, biased, not fully accurate data
- removal of context, e.g. absolutes instead of per capita
- making an argument
- downplaying or overemphasizing a risk
:::

## Do you even need a chart?

Sometimes we assume that because we have data, we should just make a chart. That
can come with some tradeoffs that we'll talk about over the course of the
semester: charts abstract data, take up space in documents, and require certain
types of literacy. They also take a lot of time and skill to make well. If
you're short on time, space, or knowhow, you might be better off using something
simple like bullet points or pullout text to briefly describe your data, rather
than making a chart that is big, unwieldy, or misleading.

For example, I analyze wage data to measure what's referred to as the wage gap.
From that, I know that Maryland workers ages 25 and up have median earnings of
\$65,586. On its own, there's no reason to make a chart.

::: {.aside}
As an aside, the reason why I always do this analysis for adults ages 25+ is to
be able to control for education.
:::

```{r}
#| label: wages1
#| fig-width: 6

library(ggplot2)
library(dplyr)

justviz::wages |>
    filter(dimension == "total", status == "all_workers") |>
    ggplot(aes(x = status, y = earn_q50)) +
    geom_col() +
    scale_y_continuous(
        expand = expansion(mult = c(0, 0.05)),
        labels = scales::label_dollar(
            accuracy = 1,
            scale = 1 / 1000,
            suffix = "k"
        )
    ) +
    labs(
        title = "Median individual earnings, Maryland adults ages 25+, 2024",
        subtitle = "This is so silly",
        x = NULL,
        y = NULL
    )
```

If I have a few more numbers I might be able to justify taking up the space:

```{r}
#| label: wages2

justviz::wages |>
    filter(dimension == "total", status != "all_workers") |>
    ggplot(aes(x = status, y = earn_q50)) +
    geom_col() +
    scale_y_continuous(
        expand = expansion(mult = c(0, 0.05)),
        labels = scales::label_dollar(
            accuracy = 1,
            scale = 1 / 1000,
            suffix = "k"
        )
    ) +
    labs(
        title = "Median individual earnings by work status, Maryland adults ages 25+, 2024",
        x = NULL,
        y = NULL
    )
```

If this is a handout or report of some sort, I'm probably still better off with
just text:

::: {.lead}
> Of Maryland workers ages 25 and up, part-time workers earn a median \$25,488,
> while full-time workers earn \$77,203.
:::

More likely, I'd round the numbers to the nearest thousand, and describe one
number in relation to the other:

::: {.lead}
> Of Maryland workers ages 25 and up, part-time workers earn a median of just
> over \$25,000. Full-time workers average 3 times as much.
:::

More numbers than this, especially with these large dollar amounts, becomes
unwieldy. For a few observations, a table works without being overwhelming:

```{r}
#| label: wages3

justviz::wages |>
    filter(dimension == "total") |>
    select(status, median_earnings = earn_q50) |>
    mutate(median_earnings = scales::label_dollar()(median_earnings)) |>
    knitr::kable(
        align = "lr",
        caption = "Median individual earnings by work status, Maryland adults ages 25+, 2024"
    )
```

We'll talk about tables as a type of data visualization later, but if I want to
make anything more than this, *and* I don't need my reader to be able to read
every single number, I'll move to a chart.

```{r}
#| label: wages4

justviz::wages |>
    filter(dimension == "by_sex") |>
    ggplot(aes(x = status, y = earn_q50, fill = sex)) +
    geom_col(width = 0.8, position = position_dodge2()) +
    scale_y_continuous(
        expand = expansion(mult = c(0, 0.05)),
        labels = scales::label_dollar(
            accuracy = 1,
            scale = 1 / 1000,
            suffix = "k"
        )
    ) +
    labs(
        title = "Median individual earnings by sex and full-time status, \nMaryland adults ages 25+, 2024",
        x = NULL,
        y = NULL
    )
```

That would be hard to cram into a short sentence, and the large gaps likely
won't be as apparent in a table, so this is now a good candidate for a chart.

Now that I've got data by both work status and sex, I might flip the story a
little. Showing income by sex could be useful, but usually when we talk about
wage gaps, it's the ratio of one group's income to another's, e.g. "women on
average make XX cents on the male dollar." If we just calculate that for the
numbers in this chart, and only want to present that, we're back down to only 3
numbers.

```{r}
#| label: wages5

ratios <- justviz::wages |>
    filter(dimension == "by_sex") |>
    tidyr::pivot_wider(
        id_cols = status,
        names_from = sex,
        values_from = earn_q50
    ) |>
    mutate(ratio_women_to_men = round(women / men, digits = 2))

ratios |>
    mutate(across(c(men, women, ratio_women_to_men), scales::label_dollar())) |>
    # select(status, ratio_women_to_men) |>
    knitr::kable(
        align = "lrrr",
        caption = "Median individual earnings by sex and full-time status, Maryland adults ages 25+, 2024"
    )
```

If my readers need to know exact dollar amounts and the ratio, a table like this
works. If all they need are ratios, you might skip the chart and table and go
for pullout text:

::: {.lead}
> Among Maryland workers ages 25 and over, women earn an average of 82 cents on
> the male dollar. Because women are more likely to work part-time, the wage gap
> is smaller between only part-time workers or only full-time workers.
:::

**Tl;dr:** Make a chart because you *should*, not because you *could*.

## Lab

The lab for this topic will be a quick check that you have all the packages you
need installed and working.
