---
title: 03. Visual encodings
execute:
  echo: false
  warning: false
  message: false
---

We'll work with a subset of the wages data in justviz and follow along some of
what's in @wilke2019, ch. 2 and 20. In [Table
2.1](https://clauswilke.com/dataviz/aesthetic-mapping.html#tab:basic-data-types),
Wilke goes over types of variables that we might find in our data and the scales
that fit each.

```{r}
#| label: setup
#| fig-height: 2

library(ggplot2)
library(dplyr)

# pull a qualitative palette from carto colors
qual_pal <- rcartocolor::carto_pal(name = "Bold")
# colorspace::swatchplot(qual_pal)
gender_pal <- qual_pal[c(3, 7)]

# pull a diverging palette but make center value gray
div_pal <- rcartocolor::carto_pal(n = 3, name = "Earth")
div_pal[2] <- "gray40"

sex_x_edu <- justviz::wages |>
    filter(dimension %in% c("by_edu", "by_sex_x_edu"), status == "full_time") |>
    select(sex, edu, earn_q20, earn_q50, earn_q80)

sex_x_edu |>
    # head() |>
    knitr::kable(caption = "Wages data by sex & education")
```

In this subset of data, we have 2 qualitative variables (sex and educational
attainment) and 3 continuous quantitative ones (20th, 50th, and 80th percentile
earnings).

On the ACS, the Census Bureau only asks gender as male or female, so there are
no gradations recorded. Educational attainment starts off quantitative: the way
it's measured is roughly the number of years of school completed (i.e.
discrete), but it could also be measured to include fractions of a year (i.e.
continuous). For simplicity's sake, however, I lumped it into broader categories
that no longer have specific identifiable numbers associated.

I also only included people with positive earnings in the past year in my
analysis, so earnings are bounded at $1.

Here's the foundation for a chart of median earnings by education.

```{r}
#| label: wages-fdn

sex_x_edu |>
    filter(sex == "total") |>
    ggplot(aes(x = edu, y = earn_q50)) +
    scale_y_continuous(breaks = seq(0, 1e6, by = 1e4))
```

So far we've got a discrete x axis for education and a continuous y axis for
median earnings. Without setting any geometries (bars, points, etc) the default
y range is just based on what's included in the data, which goes from about 40k
to somewhere above 110k (I've added extra breaks on the y-axis to see this
better).

If I plot points, this range doesn't need to change.

```{r}
#| label: wages-dot

sex_x_edu |>
    filter(sex == "total") |>
    ggplot(aes(x = edu, y = earn_q50)) +
    geom_point(size = 6) +
    scale_y_continuous(breaks = seq(0, 1e6, by = 1e4))
```

However, I want to make a bar chart. Points encode values to their *positions*,
whereas bar charts encode their data to the *length* of the bar. For a value to
correspond to the length of a bar, it needs to start at 0. By adding a column
geometry, ggplot knows to start the y-axis at 0.

```{r}
#| label: wages-bar-1

sex_x_edu |>
    filter(sex == "total") |>
    ggplot(aes(x = edu, y = earn_q50)) +
    geom_col(width = 0.8) +
    scale_y_continuous(breaks = seq(0, 1e6, by = 1e4))
```

So now we've got a discrete axis for education, a continuous axis for median
earnings, and bars whose lengths give the median earnings for each group. We
have room for more encodings:

```{r}
#| label: wages-sex-1

sex_x_edu |>
    filter(sex != "total") |>
    ggplot(aes(x = edu, y = earn_q50, fill = sex)) +
    geom_col(width = 0.8) +
    scale_fill_manual(values = gender_pal)
```

**Something's wrong,** though: by stacking the bars, we're saying that there's
some meaning to the combined lengths. That could be the case for some data (e.g.
median wage earnings in one group stacked with median income from other sources
like pension in another group could, in fact, give a sense of total income from
multiple sources). But in this case, the average woman's earnings plus the
average man's earnings doesn't mean anything. So the bars should be clustered
instead.

```{r}
#| label: wages-sex-2

sex_x_edu |>
    filter(sex != "total") |>
    ggplot(aes(x = edu, y = earn_q50, fill = sex)) +
    geom_col(width = 0.8, position = position_dodge2()) +
    scale_fill_manual(values = gender_pal) +
    scale_y_continuous(breaks = seq(0, 1e6, by = 1e4))
```

Now sex is encoded to color. Notice that the y-axis has to go a little higher
now to fit values for men with graduate degrees.

We could also include other percentiles of earnings, not just medians. The
*type* of continuous value on the y-axis will stay the same, but the *range* of
values will change again, because the 20th percentile values will be lower than
the medians and the 80th percentile values will be much higher.

Just income percentiles by education:

```{r}
#| label: wages-ptile-bar

sex_x_edu |>
    filter(sex == "total") |>
    tidyr::pivot_longer(
        -sex:-edu,
        names_to = "percentile",
        values_to = "earnings"
    ) |>
    mutate(percentile = forcats::as_factor(percentile)) |>
    ggplot(aes(x = edu, y = earnings, fill = percentile)) +
    geom_col(width = 0.8, position = position_dodge2()) +
    scale_fill_manual(values = div_pal) +
    scale_y_continuous(breaks = seq(0, 1e6, by = 1e4))
```

This is okay, but it's not super interesting. It doesn't show me anything
besides what I would expect---that people with higher educational attainment
average higher earnings, which we already knew, and that the 80th percentiles
are much higher than the 20th percentiles, which is just true by their
definition. What's more interesting are the *gaps* between values. Using points
instead of bars lets us zoom in a little better on the actual range of values
since we don't have to include 0, and it shifts what our attention is drawn to.

```{r}
#| label: wages-ptile-1

sex_x_edu |>
    filter(sex == "total") |>
    tidyr::pivot_longer(
        -sex:-edu,
        names_to = "percentile",
        values_to = "earnings"
    ) |>
    mutate(percentile = forcats::as_factor(percentile)) |>
    ggplot(aes(x = edu, y = earnings, color = percentile)) +
    geom_point(size = 6) +
    scale_color_manual(values = div_pal) +
    scale_y_continuous(breaks = seq(0, 1e6, by = 1e4))
```

Rather than just reading amounts by the ends of bars, this emphasizes the range
of values within each group. The range from 20th to 80th percentiles within each
group is implied, but we can make it explicit by adding paths behind the points.

```{r}
#| label: wages-ptile-2

sex_x_edu |>
    filter(sex == "total") |>
    tidyr::pivot_longer(
        -sex:-edu,
        names_to = "percentile",
        values_to = "earnings"
    ) |>
    mutate(percentile = forcats::as_factor(percentile)) |>
    ggplot(aes(x = edu, y = earnings, color = percentile)) +
    geom_path(color = "gray80", linewidth = 3) +
    geom_point(size = 6) +
    scale_color_manual(values = div_pal)
```

If I want the focus to be on medians, I can add size of points as another
encoding, making the medians larger and other percentiles smaller. By default,
size would be assumed to be continuous, but you can also make it discrete. (I
might do this on the fly in practice, but here I'll create a separate variable):

```{r}
#| label: wages-ptile-3

sex_x_edu |>
    filter(sex == "total") |>
    tidyr::pivot_longer(
        -sex:-edu,
        names_to = "percentile",
        values_to = "earnings"
    ) |>
    mutate(percentile = forcats::as_factor(percentile)) |>
    mutate(
        ptile_type = ifelse(percentile == "earn_q50", "median", "endpoint")
    ) |>
    ggplot(aes(
        x = edu,
        y = earnings,
        color = percentile,
        size = ptile_type,
        group = edu
    )) +
    geom_path(color = "gray80", linewidth = 3) +
    geom_point() +
    scale_color_manual(values = div_pal) +
    scale_size_manual(values = c(median = 6, endpoint = 3.5))
```

Instead of size, this could have also been something like point shape. This is
what Wilke describes as *redundant encodings*, where the percentile comes from
both color and size. I've chosen colors that should be colorblind-friendly, and
the points in this chart don't overlap, but in some situations having 2
encodings might help make things easier to read (see his examples with
scatterplots with both color and shape).

In fact, let's imagine we need a chart that can be printed in black and white.
This color palette probably won't hold up, so we can go ahead and add shape as a
backup encoding if color fails.

::: {.aside}
Call `?scale_shape` to get a cheat sheet of valid shape labels and codes. I have
to do this every time.
:::

```{r}
#| label: wages-ptile-4

sex_x_edu |>
    filter(sex == "total") |>
    tidyr::pivot_longer(
        -sex:-edu,
        names_to = "percentile",
        values_to = "earnings"
    ) |>
    mutate(percentile = forcats::as_factor(percentile)) |>
    mutate(
        ptile_type = ifelse(percentile == "earn_q50", "median", "endpoint")
    ) |>
    ggplot(aes(
        x = edu,
        y = earnings,
        color = percentile,
        fill = percentile,
        size = ptile_type,
        group = edu,
        shape = percentile
    )) +
    geom_path(color = "gray80", linewidth = 3) +
    geom_point() +
    scale_color_manual(values = div_pal, aesthetics = c("color", "fill")) +
    scale_size_manual(values = c(median = 6, endpoint = 3.5)) +
    scale_shape_manual(values = c(earn_q20 = 25, earn_q80 = 24, earn_q50 = 18))
```

There are more things we could do now, but I'll leave it here. We still have
data by sex, which I'm really interested in, but we've maxed out encodings that
we've used so far. It would be confusing to add another pair of colors or
shapes, for example, so one pair can be men and another pair could be women.
Instead, I'd likely split the plot into facets to get small multiples---we'll
talk about that later but see if you can imagine what that might look like.

## Labs

For this week's labs, you'll practice identifying the encodings in different
types of charts, then fix a chart with inappropriate encodings.

