---
title: Hacking together a legend
---

Here's a way to get a legend when your main dataset doesn't necessarily get you one: create a little dummy dataset of your aggregate values or baselines or whatever else, and assign its values to the encodings you want. Here I make a data frame of just the mean and median across all tract observations, then "tidy" it into a long-shaped data frame so I have one column with the measure type and one with the corresponding values.

```{r}
#| label: setup
library(ggplot2)
acs_tract <- justviz::acs |>
    dplyr::filter(level == "tract", !is.na(homeownership))
acs_aggs <- acs_tract |>
    dplyr::summarise(
        mean_homeownership = mean(homeownership),
        median_homeownership = median(homeownership)
    )
acs_aggs
```


```{r}
#| label: tidy
acs_aggs_tidy <- acs_aggs |>
    tidyr::pivot_longer(
        cols = dplyr::everything(),
        names_to = "measure",
        values_to = "homeownership"
    )
acs_aggs_tidy
```


Now assign measure to some aesthetics that will have legends associated with them, such as color and linetype:

```{r}
#| label: legend
ggplot(acs_tract, aes(x = homeownership)) +
    geom_density() +
    geom_vline(
        aes(xintercept = homeownership, color = measure, linetype = measure),
        data = acs_aggs_tidy
    ) +
    scale_linetype_manual(values = c("dashed", "solid"))
```

This can get kind of hacky. In fact, I went looking on Stack Overflow for questions about hacking ggplot legends (there are [7,716 questions](https://stackoverflow.com/search?q=%5Bggplot2%5D+legend+is%3Aquestion) tagged "ggplot2" about legends) and ended up [answering one myself](https://stackoverflow.com/a/80005683/5325862), albeit one based on some bad design choices.