---
title: Basic data wrangling
---

Some notes on wrangling data for visualization. We won't have time to go into a
lot of detail, but there are lots of resources available to help solve these
sorts of problems, such as:

- Relevant package docs:
  - [`dplyr`](https://dplyr.tidyverse.org/)
  - [`tidyr`](https://tidyr.tidyverse.org/)
  - [`forcats`](https://forcats.tidyverse.org/)
- [ Cheatsheets ](https://opensource.posit.co/resources/cheatsheets/) from
  RStudio, including many of the tidyverse packages
- Chapters 3 & 5 of @wickham2023
- Stack Overflow, particularly the [curated r-faq
  tag](https://stackoverflow.com/questions/tagged/r-faq)

Also, for folks who haven't yet taken GES 668, Building Spatial Datasets, see
that class's [online notes](https://bldgspatialdata.github.io/website/),
especially [Week
4](https://bldgspatialdata.github.io/website/weeks/week_04.html).

::: {.aside}
I was able to quickly get the notes for our class set up by copying the basic
template [Eli](https://github.com/elipousson) was using for Building Spatial
Datasets. Shout out to open source.
:::

Some concerns in wrangling data as they relate to our work with ggplot:

- the shape (is it a [tidy format](https://r4ds.hadley.nz/data-tidy.html), i.e.
  has observations of given types in rows and features/variables in columns)
- what's included (filtering out data we don't need to visualize)
- additional encodings (deriving new variables from existing ones)

I like to think through what I'll need in order to make the plot I want,
especially what visual encodings there will be, then work backwards from there
as needed.

## Example 1

I want to use the ACS data to make a stacked bar chart of education level for
the US, the state, and a few counties. The encodings I'll need are:

- **location** (discrete) on the **y**-axis, ordered by geographic level (nation
  / state / county)
- **value** / percent share (continuous) on the **x**-axis
- **education** (discrete ordinal) as **fill**, ordered by level (less than high
  school / high school only / some college or associate's / bachelors / graduate
  degree)

Note that for a stacked bar chart, each segment will correspond to the share of
adults in each location with each level of education, e.g. there will be one
segment for the share of adults in Baltimore city with a graduate degree; its y
position will be in a row for the city, its length will correspond to the value,
and its position within the bar will depend on the education level relative to
the other education levels (should be the last level).

First, I'll subset the data for just the rows (US, MD, selected counties) and
columns (level, name, all the education-related variables) I need. With most
`dplyr` functions, you can use bare column names (i.e. don't need quotation
marks). This is part of a more complicated programming paradigm that I can
barely explain, certainly not in these notes.

```{r}
#| label: setup

library(ggplot2)

# grab a sequential color palette
seq_pal <- RColorBrewer::brewer.pal(n = 5, name = "YlGnBu")

edu <- justviz::acs |>
    dplyr::filter(
        name %in%
            c(
                "United States",
                "Maryland",
                "Baltimore city",
                "Baltimore County",
                "Anne Arundel County",
                "Howard County"
            )
    ) |>
    dplyr::select(
        level,
        name,
        less_than_high_school,
        high_school_grad,
        some_college_or_aa,
        bachelors,
        grad_degree
    )

edu
```

In order to have education level mapped to bar fill and percentage of adults
with each education level by location mapped to the length of a bar segment,
I'll need one variable to assign to fill and one variable to assign to x. That
means I need to reshape the data from a *wide* format to a *long* format, i.e. I
need to tidy it. `tidyr::pivot_longer` will take us from a wider data frame to a
longer one. (Guess what `tidyr::pivot_wider` does.) It has some complicated
arguments but we'll keep it simple: which columns will be reshaped into a single
column, what will that column be called, and what will those columns' values be
called.

::: {.aside}
Older code will use `tidyr::gather` and `tidyr::spread`; these still work, but
have been deprecated a few years now.
:::

```{r}
#| label: tidy

edu_tidy <- edu |>
    # there are other ways to select columns---see tidyselect docs---but it's good to start out explicit
    tidyr::pivot_longer(
        cols = c(
            less_than_high_school,
            high_school_grad,
            some_college_or_aa,
            bachelors,
            grad_degree
        ),
        names_to = "education",
        values_to = "share"
    )

edu_tidy
```

Now each row is a unique combination of geographic level, location, education
level, and percentage. I could plot it like this to get started:

```{r}
#| label: plot-1

edu_tidy |>
    ggplot(aes(x = share, y = name, fill = education)) +
    geom_col(width = 0.8, position = position_fill()) +
    scale_fill_manual(values = seq_pal)
```

Pay attention to the orders of our categorical variables: location (y-axis) is
in alphabetical order starting at the bottom, and education level is in
alphabetical order from the top of the bar to the bottom. You could keep
location in alphabetical order and just flip it (A on top, Z on bottom); it's
not wrong. In my work, however, it's usually more correct to order first by
geographic level, then alphabetically within geographic levels.

The most surefire way to get categorical variables in the order you want is to
make them into factors, and `forcats` is a great package for helping with that.

By default, R orders factors alphabetically. We can then reverse the order of
levels with `forcats::fct_rev`.

```{r}
#| label: plot-factor-rev

edu_tidy |>
    dplyr::mutate(name = factor(name) |> forcats::fct_rev()) |>
    ggplot(aes(x = share, y = name, fill = education)) +
    geom_col(width = 0.8, position = position_fill()) +
    scale_fill_manual(values = seq_pal)
```

If we want to order locations by geographic level, we have a few options.
`forcats::as_factor` will convert the variable to a factor, with levels ordered
as they appear. I set the data up to be ordered by geographic level, so this is
actually all you need here because I was looking out for you. But if that
weren't the case, you could make geographic level a factor in a logical order,
then make location name a factor, then arrange the data frame by geographic
level and name, then make location name a factor with that order. Here's the
more complicated version of that so you have it as reference:

```{r}
#| label: plot-factor-reorder

edu_tidy |>
    dplyr::mutate(level = forcats::as_factor(level)) |>
    dplyr::arrange(level, name) |>
    dplyr::mutate(name = forcats::as_factor(name) |> forcats::fct_rev()) |>
    ggplot(aes(x = share, y = name, fill = education)) +
    geom_col(width = 0.8, position = position_fill()) +
    scale_fill_manual(values = seq_pal)
```

While the order of locations didn't have a single right answer, the order of
education levels does. One is by definition higher than another. I put the
columns in order when I built the data frame (again, looking out for you), and
when `tidyr::pivot_longer` reshapes the data, it keeps education in the order it
encounters their columns, so we can just use that ordering and make education a
factor:

```{r}
#| label: plot-ordinal

edu_tidy |>
    dplyr::mutate(level = forcats::as_factor(level)) |>
    dplyr::arrange(level, name) |>
    dplyr::mutate(name = forcats::as_factor(name) |> forcats::fct_rev()) |>
    dplyr::mutate(education = forcats::as_factor(education)) |>
    ggplot(aes(x = share, y = name, fill = education)) +
    geom_col(width = 0.8, position = position_fill()) +
    scale_fill_manual(values = seq_pal)
```

(If you didn't have a professor who set things like this up to make it easier on
you, you could also order the education levels manually with
`forcats::fct_relevel`.)

Lastly, we'll clean up the levels of the education factor. Right now they have
values that can work as syntactically valid column names---that's where they
came from, after all. I like the `snakecase` package for starting my string
cleaning like this, then doing manual clean-up as needed. `forcats::fct_relabel`
takes a function to apply (I love functional programming but we won't get to do
much with it), and `forcats::fct_recode` replaces levels manually.

```{r}
#| label: plot-strings

edu_tidy |>
    dplyr::mutate(level = forcats::as_factor(level)) |>
    dplyr::arrange(level, name) |>
    dplyr::mutate(name = forcats::as_factor(name) |> forcats::fct_rev()) |>
    dplyr::mutate(
        education = forcats::as_factor(education) |>
            forcats::fct_relabel(snakecase::to_sentence_case) |>
            forcats::fct_recode(
                "Some college or associate's" = "Some college or aa",
                "Bachelor's" = "Bachelors",
                "Graduate degree" = "Grad degree"
            )
    ) |>
    ggplot(aes(x = share, y = name, fill = education)) +
    # add reverse = TRUE to order from bottom of bar to top
    geom_col(width = 0.8, position = position_fill(reverse = TRUE)) +
    scale_fill_manual(values = seq_pal)
```

## Example 2

Next I'll make changes to the CDC dataset to make a table of multiple variables.
As it stands, the data is already tidy, but for a table I'd print, I actually
want it untidy. That is, I'll want each health condition as a column, and each
location as a row.

::: {.callout-warning}
FYI You'll need `justviz` version 0.2.6 for this, because I just found a bug in
how the CDC data labeled Baltimore city vs Baltimore County (they were both just
Baltimore) while working on this!
:::

First I'll pull together a subset of the data for the country, state, Baltimore
city, and Baltimore County, and just a few indicators I'm interested in. These
would all make sense to go in a table together to study access to health care.

```{r}
#| label: cdc-setup

health_risk <- justviz::cdc |>
    dplyr::filter(
        location %in% c("US", "Maryland", "Baltimore city", "Baltimore County")
    ) |>
    dplyr::filter(
        indicator %in%
            c(
                "Annual checkup",
                "Health insurance",
                "Dental visit",
                "Mobility disability"
            )
    ) |>
    # don't need population column
    dplyr::select(-pop)

health_risk
```

A few things I notice:

- In the chart, I had United States instead of US, and I want that to match here
- Locations should be in geographic order again (US, MD, counties
  alphabetically)
- Some indicator labels are unclear, but because I read the [data
  documentation](https://data.cdc.gov/d/swc5-untb), I know how to relabel them
- Values will be printed as percentages, but right now are written as though
  they have a denominator of 100 already (i.e. 12 instead of 0.12 for 12%). I
  can decide what to do for formatting numbers.
- ~~Indicators are from either 2022 or 2023, and I want to note the year in the column names in my table~~
  I'll do this as a second example

```{r}
#| label: cdc-label

# same type of formatter we've defined before, but give it 1 decimal place
percent <- scales::label_percent(accuracy = 0.1)

health_relabeled <- health_risk |>
    dplyr::arrange(level) |>
    dplyr::mutate(location = forcats::as_factor(location)) |>
    dplyr::mutate(
        indicator = forcats::as_factor(indicator) |>
            forcats::fct_recode("No health insurance" = "Health insurance") |>
            # reorder to start with access, then outcomes
            forcats::fct_relevel(
                "No health insurance",
                "Mobility disability",
                "Annual checkup",
                "Dental visit"
            )
    ) |>
    # convert back to decimal to use a percent formatter
    dplyr::mutate(value = percent(value / 100)) |>
    # calling arrange again confirms I have factors in the order I want
    dplyr::arrange(location, indicator)

health_relabeled
```

Now that the data is formatted how I want, I can reshape it from long to wide.
The new columns will be based on unique items in the indicator column, and their
corresponding values will be what's in the value column.

```{r}
#| label: cdc-pivot

health_wide <- health_relabeled |>
    # think of the id_cols argument as an anchor while you reshape the data
    tidyr::pivot_wider(
        id_cols = c(level, location),
        names_from = indicator,
        values_from = value
    ) |>
    dplyr::rename(Level = level, Location = location)

health_wide
```

Since there are 2 different years in this data, I might want to include the year
of each indicator in its column label. I can use more than one column to create
names, and supply a separator to go between them:

```{r}
#| label: cdc-pivot2

health_relabeled |>
    tidyr::pivot_wider(
        id_cols = c(level, location),
        names_from = c(indicator, year),
        values_from = value,
        names_sep = ", "
    ) |>
    dplyr::rename(Level = level, Location = location)
```

## Read the docs

Realistically, you'll probably run into other data wrangling needs in this
course. I can't predict them all, but these are some of the basic ones you'll
most likely encounter, and we can work through any others as they come. Also, as
always, **READ THE DOCS**.
